\DiaryEntry{Cyclic (finite) Groups}{2017-09-01}{Algebra}

This topic keeps coming up again and again and this entry will try to collect all information once and forever.


\subsection{Definitions}

The order of a group $G$ is denoted as $|G|$ and is the number of group elements. The order $|x|$ of a group element $x$ is the smallest integer $n$ for which $x^n = e$ holds. Note that the order of each group element can be different, but is in the range of $1$ (the identity element) and $|G|$ (there simply aren't more group elements).

Let us write down the sequence of $x^i$:

\be\label{cyclic_groups_1:eq}
x^0 = e, x^1 = x, x^2, x^3, \ldots, x^{n-1}, x^n = e
\ee

For a finite group ($|G| = n < \infty$), the elements of this sequence are all different: Assume that $0 \leq m < a < b < n$ and assume $x^a = x^b$, then $x^{b-a}=1$. With the before assumption, $b-a < n$ and this contradicts the definition of the order of $x$.

If a group element $g$ has the same order as the group order, it is called a generator for the group and we write $G = \langle g \rangle$. As can be seen in the sequence \eqref{cyclic_groups_1:eq}, in this case $g^i$ will run through all group elements; i.e. generate the group. If the order of an element is less than $n$, then above sequence will not cover all elements of the group.

Intuitively, since all elements in above sequence \eqref{cyclic_groups_1:eq} are different, the group contains $N$ elements, and the order of a group element is also $N$, then the sequence must generate all group elements. If the group element order is smaller, then the sequence does not generate all group elements. We will find out further below, that in this case a cyclic subgroup is created.


\subsection{Example}

Consider the group $\mZ_5^\star$. Its order is $|G| = 4$ and the group elements are all those numbers being coprime with $5$; i.e. $\mZ_5^\star= \{1,2,3,4\}$ (see also \hyperref[2017-05-05:entry]{here}). Let's write down the sequence \eqref{cyclic_groups_1:eq} for each element:

\be\label{cyclic_groups_2:eq}
\begin{array}{c|ccccc}
	     i      & 0 & 1 & 2 & 3 & 4 \\ \hline
   	1^i \bmod 5 & 1 & 1 & 1 & 1 & 1 \\
	2^i \bmod 5 & 1 & 2 & 4 & 3 & 1 \\
	3^i \bmod 5 & 1 & 3 & 4 & 2 & 1 \\
	4^i \bmod 5 & 1 & 4 & 1 & 4 & 1
\end{array}
\ee

In Julia, this can be achieved via

\begin{verbatim}
[mod(2^i,5) for i in 1:4]
\end{verbatim}

We can see that both $2, 3$ have order $4$, their powers run through all group elements and are therefore generators for $\mZ_5^\star$. The element $4$ has order $2$, its powers do not run through all group elements and therefore the element is not a generator for the group.

\begin{theorem}
	If $G$ is a group, let $x \in G$ and let $a$ be an integer unequal zero.

	We have for $|x| = n < \infty$: $|x^a| = n / \gcd(n,a)$.

	Assume $G = \langle g \rangle$ and $|x| = n < \infty$. Then $G = \langle x^a \rangle$ iff $\gcd(a,n) = 1$. In particular, the number of generators of $G$ is $\phi(n)$.

\end{theorem}

Note that this theorem can only generate new (additional) generators ($x^a$) when one generator ($x$) is known.

In our example, we have $|G| = 4$ and $\phi(4) = 2$. This is in line with the observations above, where we identified two generators, namely $2$ and $3$. Let's see if we can obtain other generators out of these two: The smallest number being coprime with $n$ is $3$ ($\gcd(3,4) = 1$). Using $2^3 = 8 \equiv 3 \bmod 5$ which gives nothing new. Using $3^3 = 27 \equiv 2 \bmod 5$ which is also nothing new.

Elements of the form $x^2$ are not generators, as $\gcd(2,4) \neq 1$. For example, $2^2 = 4$ is not a generator.

Finally, we can provide the subgroup structure of a cyclic subgroup as follows:

\begin{theorem}
	Let $G = \langle g \rangle$ with $|G| = n < \infty$.
	\begin{itemize}
		\item Every subgroup of $G$ is cyclic. If $K \leq G$, then either $K = \{1\}$ or $K = \langle x^d \rangle$, where $d$ is the smallest positive integer such that $x^d \in K$.
		\item For each positive integer $a$ dividing $n$, there is a unique subgroup of $G$ with order $a$. This subgroup is the cyclic subgroup $\langle x^d \rangle$, where $d = n / a$.
	\end{itemize}
\end{theorem}

In the example, $2$ is a divisor of $|G| = 4$, so there must exist a cyclic subgroup of order $2$. Let us try the generator $2$ first. We have $H_1 = \langle 2^2 \rangle =  \langle 4 \rangle = \{1,4\}$. It is straightforward to check that this is a subgroup of $\mZ_5^\star$. Using the other generator, $3$, we have $H_2 = \langle 3^2 = 9 \equiv 4 \bmod 5$ which is the same subgroup as $H_1$. This is no coincidence; the theorem above says that there is a \emph{unique} subgroup of order $a$.


\subsection{Example $\mZ_7^\star$}

The group contains all integers which are coprime with $7$ and with $7$ being a prime this are the following $\mZ_7^\star =\{1,2,3,4,5,6\}$. The group order is therefore $|\mZ_7^\star| = 6$ and the group contains $\phi(6) = 2$ generators. Let's see what they are by writing down the sequence for each group element

\bee
\begin{array}{c|cccccccc}
	     i      & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\ \hline
	1^i \bmod 7 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
	2^i \bmod 7 & 1 & 2 & 4 & 1 & 2 & 4 & 1 \\
	3^i \bmod 7 & 1 & 3 & 2 & 6 & 4 & 5 & 1 \\
	4^i \bmod 7 & 1 & 4 & 2 & 1 & 4 & 2 & 1 \\
	5^i \bmod 7 & 1 & 5 & 4 & 6 & 2 & 3 & 1 \\
	6^i \bmod 7 & 1 & 6 & 1 & 6 & 1 & 6 & 1 \\
\end{array}
\eee

The two generators are $3$ and $5$. With group order $6$, the group contains subgroups of order $2$ and $3$, respectively. The order-$2$ subgroup is $\langle 6 \rangle = \{1, 6\}$, the order-$3$ subgroup is $\langle 2 \rangle =\{1,2,4\}$. Another order-$3$ subgroup is  $\langle 4 \rangle =\{1,2,4\}$ which is isomorphic to $\langle 2 \rangle$.

\subsection{Example $\mZ_{11}^\star$}

Another example (with a hopefully more interesting subgroup structure) is

\bee
\mZ_{11}^\star= \{1,2,3,4,5,6,7,8,9,10\}
\eee

Having $10$ elements, the group contains $\phi(10) = 4$ generators which we find to be $2,6,7,8$ (experimentally). The other elements are generators for the following subgroups

\begin{align*}
\langle 3 \rangle & = \{1,3,4,5,9\}  \\
\langle 10 \rangle & = \{1,10\}
\end{align*}

Their order is $2$ and $5$, the only numbers $10$ is divisible by.

\subsection{Example $\mZ_{13}^\star$}

As a final example consider this group. With $\phi(13) = 12$ the group order is $12$, and has $\phi(12) = 4$ generators. Since $12$ is divisible by $2,3,4,6$, we have subgroups of these orders.

The generators are $2, 6, 7, 11$. The other elements are generators for the following subgroups

\begin{align*}
\langle 3 \rangle = \langle 9 \rangle & = \{1,3,9\}  \\
\langle 4 \rangle = \langle 10 \rangle & = \{1,3,4,9,10,12\} \\
\langle 5 \rangle = \langle 8 \rangle & = \{1,5,8,12\} \\
\langle 12 \rangle & = \{1,12\}
\end{align*}

\subsection{Final Observation and Outlook}

In order to obtain cyclic subgroups, one takes a singleton subset $\{x\}$ of the group $G$. In order to obtain a subgroup, one has to \emph{close} the set under group operation and taking inverses. This closing is achieved by adding all integral powers of $x$ (since $x^n = e$, this will eventually include the neutral group element as well); the result is a subgroup. This subgroup is the smallest subgroup of $G$ which contains $x$. One can extend this closure operation and consider arbitrary subsets of $G$ instead of $\{x\}$.