\DiaryEntry{Knapsack Problem}{2022-01-17}{}

We have a set of $N$ different item types, each type has a weight $w_i$ and a value $v_i$. We want to pack $x_i$ items ($x_i \in \mN$; further constraints see below) of each type into a knapsack with maximum weight capacity $W$ so that the overall value $m[W]$ is maximized,

\bee
m[W] = \max_{x_i} \sum_{i=1}^N v_i w_i
\eee

with the constraints

\bee
  \sum_{i=1}^N w_i x_i \leq W
\eee

The weights must be strictly positive ($w_i > 0$), otherwise any solution could be improved (unboundedly) by adding an infinite number of items with zero or negative weight.

There are several possibilites for the allowed values of the $x_i$; the \emph{0-1 Knapsack Problem} allows at most one item of each type; i.e. $x_i \in \{0,1\}$. In a \emph{bounded Knapsack problem}, the integers $x_i$ are restricted to a maximum number of each item type denoted by the integer $c$. Finally, the \emph{unbounded Knapsack problem} has no bounds on the integers $x_i$.


\subsection{Unbounded Knapsack Problem}

We solve the Knapsack problem in a recursive manner (using a dynamic programming approach); i.e. we solve it for a knapsack with capacity $W = 0$, then for a knapsack with capacity $W=1$ and so on. We define the function

\bee
m[w'] = \max_{x_i} \sum_{i=1}^N v_i w_i
\eee

with the constraint 

\bee
\sum_{i=1}^N w_i x_i \leq w'
\eee

This function has the following characteristics,

\bee
m[0] = 0
\eee

as we can not fit any items in a knapsack having empty capacity, $W = 0$. A second characteristic is

\bee
m[w] = \max(v_1 + m[w - w_1], v_2 + m[w - w_2], cdots, v_N + m[w - w_N]), \quad w_i \leq w
\eee

This can be explained as follows: Assume we want to fill a knapsack with capacity $w$ and think which item we put in next. We can choose any item $i$ out of the $N$ types and this will add a value of $v_i$ to the knapsack. However, this item has weight $w_i$ so the value $v_i$ is added to the knapsack with weight $w - w_i$ in order not to exceed capacity. We chose the item which creates the knapsack with maximum value. In the end, we arrive at $m[W]$ which holds the solution to the problem. If we additionally keep track of which item $i$ we add in each step (i.e. the $\arg \min$), then we also know which items the knapsack contains.

\paragraph{Example.}  We have $N = 2$ different items with weights $w_1 = 2, w_2 = 3$ and values $v_1 = 3, v_2 = 5$, respectively. As outlined above, we start with $m[0] = 0$. For $m[1]$ we observe that the item with the smallest weight has weight $2$, therefore we cannot add anything to the knapsack and $m[1] = 0$. In our formula, we see that there are no terms to take the maximum over and therefore

\bee
m[1] = \max_{w_i \leq 1}(v_i + m[w - w_i]) = 0
\eee

The next one is more interesting, we have

\bee
m[2] = \max_{w_i \leq 2} (v_i + m[2-w_i]) = \max(v_1 + m[2-w_1]) = \max(3 + m[0]) = \max(3) = 3
\eee

which was realized by adding item $1$ to the knapsack. Since the second item has weight $w_2 = 3 > 2$, it was not considered in the maximization. Next we consider $w=3$ and obtain

\bee
m[3] = \max_{w_i \leq 3} (v_i + m[2-w_i]) = \max(3 + m[1], 5 + m[0]) = \max(3, 5) = 5
\eee

which was realized by adding item $2$ to the knapsack. We continue with $m=4$ and get

\bee
m[4] = \max_{w_i \leq 4} (v_i + m[2-w_i]) = \max(v_1 + m[2], v_2 + m[1]) = \max(3 + 3, 5 + 0) = 6
\eee

We have added one item $1$ to the knapsack; however, the ``starting point'' was the knapsack $m[2]$ which already held one item $1$. Therefore $m[4]$ is realized by a knapsack with \emph{two items $1$}. This is typical for such dynamic programming solutions that a small increase in the function argument (going from $m[3]$ to $m[4]$) causes a different solution (similar with the Viterbi algorithm).


\subsection{Binary Knapsack Problem}

A similar dynamic programming solution for the 0-1 Knapsack Problem also runs in pseudo-polynomial time. We define $m[i,w]$  to be the maximum value that can be attained with weight $\leq w$ using only the first $i$ item types. We can define $m[i,w]$ recursively with starting point

\bee
m[0,w]  = 0
\eee

For values $i > 0$, we have two cases: (i) The new item is heavier than the current weight limit and we therefore cannot add it. In this case we have $m[i,w] = m[i-1,w]$. (ii) The new item is allowed from the weight limit. In this case we need to check whether replacing item $i-1$ by item $i$ increases value, $m[i,w] = \max(m[i-1,w], m[i-1,w-w_i]+v_i)$. Combining this, we have

\bee
m[i,w] = \begin{cases} m[i-1,w] \quad w_i > w \\
  \max(m[i-1,w], m[i-1,w-w_i]+v_i) \quad w_i \leq w \end{cases}
\eee

The solution can then be found by calculating $m[n,W]$. To do this efficiently, we can use a table to store previous computations and knapsack content. Be aware, that there is an upper value limit $\sum w_i$ we can get at most into a knapsack (as we are allowed to take at most one of each item type). So for the algorithm to make sense, we need $\sum w_i \geq W$

\paragraph{Example.} We have three different items with weights $w_1 = 2, w_2 = 3, w_3 = 5$ and values $v_1 = 3, v_2 = 5, v_3 = 7$. The weight limit is $W = 5$. We start with $i = 1$ and calculate $m[1,1]$. Since $w_i = 2 > w = 1$, the first item does not fit into the knapsack with capacity $w = 1$. This is reflected by the first condition and therefore $m[1,1] = m[0,1] = 0$. We continue with $m[1,2]$. Now the first item fits into the knapsack with capacity $w = 2 \leq w_1 = 2$ and therefore we use the second condition,

\bee
m[1,2] = \max(m[0,2], m[0,0]+  v_1) = 3
\eee

Here we choose to add the item one to the knapsack. The further values $m[1,3], m[1,4]$ and $m[1,5]$ do not change anything: We are only allowed to use item one (as $i = 1$) and of course this item will fit into the knapsack with higher capacity.

More interesting are the cases for $i = 2$. First consider $m[2,1]$ and since $w_2 > w$, we have $m[2,1] = m[1,1] = 0$. The same holds for $m[2,2] = m[1,2] = 3$ with the knapsack holding item one. Next is

\bee
m[2,3] = \max( m[1,2], m[1,0] + 5) = \max(3, 0 + 5) = 5
\eee

and we have decided to add the second item to the knapsack. For $m[2,4]$ we have

\bee
m[2,4] = \max( m[1,4], m[1, 0] + 5) = \max(3,5) = 5
\eee

where we again have chosen the second item. Finally, we calculate $m[2,5]$ as

\bee
m[2,5] = \max( m(1,4), m[1, 2] + 5) = \max( 3, 3 + 5) = 8
\eee

which we achieve by putting item one (cf $m[1,2]$) and item two into the knapsack.

Now to $i = 3$ and let's speed things up a bit. With the same arguments as before $m[3,1] = 0, m[3,2] = 3, m[3,3] = 5, m[3,4] = 5$; i.e. when the knapsack capacity allows it, we replace item one with item two. When calculating $m[3,5]$ we have

\bee
m[3,5] = \max( m[2,5], m[2,0] + 7) = \max(8, 7) = 8
\eee

where we keep the content of $m[2,5]$; i.e. item one and item two. If the value of the third item were higher; e.g. $v_3 = 10$, then we would choose $m[2,0]$ (which is empty) and add item three to it.

\paragraph{Other interesting things.} There exist so called \emph{dominance relations} whih can simplify a Knapsack problem. If two items $i, j$ have the same weight $w_i=w_j$, but one has a higher value $v_i > v_j$, then we can exclude item $j$ from the problem. In a simialr spirit, if we have a set $J$ of items with the same weight as an item $i$; i.e. $\sum_{k \in J} w_k = wi$ but smaller value, $\sum_{k \in J} w_k < v_i$, we can exclude the items from $J$ from the problem.

\todo{The available capacity cannot always be filled up; there will be some slack between $\sum w_i x_i$ and $w$. How is this slack distributed for e.g. random weights and values?}


\subsubsection{Suboptimal Solutions}

\paragraph{Greedy Algorithm.} An intuitive approach is to assign a \emph{profit to weight ratio} $e_i = p_j / w_j$ to every item. We next order the items with decreasing ratio and start adding the items with highest ration to the knapsack if the weigh constrinat allows for it.

\begin{verbatim}
wG = 0
zG = 0
for i = 0:N
   if wG + wj <= w
      xj = 1
      wG = wG + wj
      zG = zG + vj
   end
end
\end{verbatim}

This algorithm is not optimal; as example, we consider the following values for weight and values,

\begin{verbatim}
wvec = [2, 3, 6, 7, 5, 9, 4]
vvec = [6, 5, 8, 9, 6, 7, 3]
\end{verbatim}

and want to fill a knapsack with capacity $W = 9$. The greedy solution starts by adding the first two items (as they have the best profit to weight raito) and than there is only space for the last item left (which has the worst profit to weight ratio). This yields a total value of $14$.

The optimal algorithm, however, takes the first and fourth element and achieves a value of $15$.

\paragraph{Linear Programming Relaxation.} The NP complexity of the 0-1-Knapsack problem comes from the fact that the $x_i$ are binary variables. If we relax this contraint, we get a linear program which can be solved much more easily. We have

\bee
m_{rel}[W] = \max_{x_i} \sum_{i=1}^N v_i w_i
\eee

with the constraints

\begin{align*}
  & \sum_{i=1}^N w_i x_i \leq W \\
  & x \in \mR \\
  & 0 \leq x_i \leq 1
\end{align*}

This relaxation causes a potentially better value for the knapsack value; i.e. $m_{rel}[W] \geq m[W]$. This is the case when we cannot completely fill the knapsack with items. In the example above, we can fill the knapsack completely (by choosing the first two items and $w_1 + w_2 = 5 = W$). Therefore, the linear programming relaxation does not yield an increased value. However, if we use a larger knapsack with $W = 6$, then we cannot use the additional space as our smallest weight is $w_1 = 2$ and does not fit. Therefore, $m[6] = m[5] = 8$. If we solve this problem using linear programming we obtain $x_1 = x_2 = 1$ and $x_3 = 0.2$ yielding $m_{rel} = 9.4$ which fills the knapsack completely; i.e. $\sum m_i x_i = 6$. Since the first two items have already been included completely, the optimizer ``decided'' to include the third item.

Note that this is not the best choice in terms of profit to weight ratio. Had we used constraint as follows,

\begin{align*}
  & \sum_{i=1}^N w_i x_i \leq W \\
  & x \in \mR \\
  & 0 \leq x_i
\end{align*}

then the optimizer chooses $x_1 = x_3 = 0, x_2 = 2$ as the second item has the biggest profit to weight ratio. This is the unbounded Knapsack problem with \emph{real} coefficients.

\todo{Show that the solution to the linear program can be obtained by a greedy algorithm.}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "journal"
%%% End:
